---
title: "TSAR Project Assignment Part 3 - Group 3"
subtitle: "Times Series Forecasting"
## title-block-banner: true
author:
  - name: Arpad Broglie
    email: arpad.broglie@students.fhnw.ch
    corresponding: true
  - name: Teodor Glisic
    email: teodor.glisic@students.fhnw.ch
    corresponding: true
date: 06/04/2026
date-format: "YYYY"
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    self-contained: true
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
    papersize: a4
    syntax-highlighting: idiomatic
editor: visual
execute: 
  warning: false
---

# Preliminaries

```{r}
# Install needed libraries to use this report
if (!require(dplyr)) install.packages("dplyr")
if (!require(tidyr)) install.packages("tidyr")
if (!require(ggplot2)) install.packages("ggplot2")
if (!require(fpp3)) install.packages("fpp3")

#Load needed libraries
library(dplyr)
library(ggplot2)
library(tidyr)
library(fpp3)


# Read already filtered time series (tsibble) 
# with Series ID CEU1000000001 (Mining and Logging)
mySeries <- readRDS("us_employment_filtered.rds")
```

## Overview of tsibble

```{r}
# Display the first few rows of the tsibble
head(mySeries)

# Inspect the structure of the tsibble (index, keys, column data types)
str(mySeries)

# Get a summary of the columns (distribution, range, missing values)
summary(mySeries)
```

# Model 1: Mean Method

## Step 1: Extracting the Training Set

```{r}
# Extracting number of rows for training set with 80/20 split
overall_count <- (nrow(mySeries))
training_row_count <- floor((nrow(mySeries) * 0.8))
train <- mySeries %>% slice(1:training_row_count)
```

## Step 2: Train 1st Benchmark Model

```{r}
# Mean method used
mean_employment_fit <- train %>%
  model(
  Mean = MEAN(Employed)
)

# View first impression of the fitted values
augment(mean_employment_fit)
```

## Step 3: Evaluate the Model Fit of 1st Benchmark Model

### Task A

```{r}
# Plot fitted values (predictions on training set) against actual values over time
augment(mean_employment_fit) %>% 
  autoplot(.fitted, colour="blue", linetype="dashed") +
  autolayer(train, Employed) +
  labs(y= "US Employment", title="Fitted and true values for monthly US employment rates (Mean method)")
```

#### Answering questions to 1st Benchmark Model

Are the fitted values close to the actual values?

Are they systematically too high or too low?

If yes, where does it come from? (Consider how your benchmark model works!)

### Task B

```{r}
# Create a scatter plot of actual values vs. fitted values with an identity line (slope=1)
augment(mean_employment_fit) %>%
  ggplot(aes(x=Employed, y=.fitted)) + 
  geom_point() + 
  geom_abline(slope = 1, intercept = 0, linetype=2) + 
  labs(title = "Actual vs. Fitted (Mean Model)", x="Actual", y="Fitted")
           
```

#### Answering questions to 1st Benchmark Model

Are the points close to the identity line?

Are they systematically too high or too low?

If yes, where does it come from? (Relate your answer to what you saw in the time plot.)

### Task C

```{r}
# Generate diagnostic plots for the model residuals (time plot, ACF plot, histogram)
mean_employment_fit %>% gg_tsresiduals()
```

#### Answering questions to 1st Benchmark Model

Are the residuals auto-correlated? How do you decide that based on the plots?

Do the residuals have zero mean? How do you decide that based on the plots?

What do these results tell you about your model?

### Task D

```{r}
# Perform a Ljung-Box test for residual autocorrelation with 24 lags (2 * seasonal period)
augment(mean_employment_fit) %>% 
  features(.innov, ljung_box, lag=24)
```

#### Answering questions to 1st Benchmark Model

Does the test result support your conclusions from (c)?

How do you conclude that from the test result?

```{r}
# Calculate the mean of the residuals of the fitted model
residual_mean <- mean(
  augment(mean_employment_fit)$.resid, 
  na.rm = TRUE)
residual_mean
```

#### Answering questions to 1st Benchmark Model

Does the result support your conclusions from (c)?

## Step 4: Evaluate the Point Forecast Accuracy of your 1st Benchmark Model

```{r}
# Make a forecast and save the result
fc_mean <- mean_employment_fit %>% 
  forecast(h = (nrow(mySeries) - nrow(train)))

# Create a time plot 'Actual vs. Forecast'
fc_mean %>% 
  autoplot(mySeries) +
  labs(title="US Employment in Mining and Logging - Actual vs. Forecast (Mean Model)")

# Calculate the forecast accuracy metrics
fc_mean %>% 
  accuracy(mySeries)
```

#### Answering questions to 1st Benchmark Model

Interpret the **RMSE** (Root Mean Squared Error) in max. 3 sentences:

Interpret the **MAPE** (Mean Absolute Percentage Error) in max. 3 sentences:

## Step 5: Evaluate the Point Forecast Uncertainty of Your 1st Benchmark Model

```{r}
# Residual diagnostics for checking homoscedasticity and normality
mean_employment_fit %>% gg_tsresiduals()

# Display prediction interval boundaries
fc_mean %>% 
  hilo()
```

#### Answering questions to 1st Benchmark Model

Are the residuals homoscedastic? (That is, do they have constant variance?)

Are the residuals normally distributed?

Does it make sense to include these prediction intervals in your model evaluation? Why/why not?

# Model 2: Seasonal naive method with drift

## Step 1: Extracting the Training Set

```{r}
# Extracting number of rows for training set with 80/20 split
overall_count <- (nrow(mySeries))
training_row_count <- floor((nrow(mySeries) * 0.8))
train <- mySeries %>% slice(1:training_row_count)
```

## Step 2: Train 2nd Benchmark Model

```{r}
# Train the model with the seasonal naive drift method
naive_drift_employment_fit <- train %>% model (
    SNaive_Drift = SNAIVE(Employed ~ drift())
)

# View the trained data
augment(naive_drift_employment_fit)
```

## Step 3: Evaluate the Model Fit of Your 2nd Benchmark Model

### Task A

```{r}
# Plot actual training data against the model's fitted values over time
augment(naive_drift_employment_fit) %>% 
  autoplot(.fitted, colour="red", linetype="dashed") +
  autolayer(train, Employed) +
  labs(y= "US Employment", title="Fitted and true values for monthly US employment rates (Seasonal Naive with Drift method)")
```

#### Answering questions to 2nd Benchmark Model

Are the fitted values close to the actual values?

Are they systematically too high or too low?

If yes, where does it come from? (Consider how your benchmark model works!)

### Task B

```{r}
# Plot actual vs. fitted values to check deviation from the identity line (slope=1)
augment(naive_drift_employment_fit) %>% 
  ggplot(aes(x=Employed, y=.fitted)) + 
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype=2) +
  labs(title = "Actual vs. Fitted (Seasonal Naive with Drift Model)", x="Actual", y="Fitted")
```

#### Answering questions to 2nd Benchmark Model

Are the points close to the identity line?

Are they systematically too high or too low?

If yes, where does it come from? (Relate your answer to what you saw in the time plot.)

### Task C

```{r}
# View residual diagnostic plots (time plot, ACF plot, histogram) to assess model fit
naive_drift_employment_fit %>% gg_tsresiduals()
```

#### Answering questions to 2nd Benchmark Model

Are the residuals auto-correlated? How do you decide that based on the plots?

Do the residuals have zero mean? How do you decide that based on the plots?

What do these results tell you about your model?

### Task D

```{r}
# Conduct a formal Ljung-Box test for autocorrelation in the residuals (lag = 2 * seasonal period)
augment(naive_drift_employment_fit) %>% 
  features(
  .innov,
  ljung_box,
  lag = 24
)
```

#### Answering questions to 2nd Benchmark Model

Does the test result support your conclusions from (c)?

How do you conclude that from the test result?

```{r}
# Calculate the mean of the residuals to check for forecast bias
residual_mean_naive_drift <- mean(
  augment(naive_drift_employment_fit)$.resid, 
  na.rm = TRUE)
residual_mean_naive_drift
```

#### Answering questions to 2nd Benchmark Model

Does the result support your conclusions from (c)?

## Step 4: Evaluate the Point Forecast Accuracy of your 2nd Benchmark Model

```{r}
# Generate forecasts for the test set period
fc_naive_drift <- naive_drift_employment_fit %>% 
  forecast(h=(overall_count - training_row_count))

# Plot the forecast values with prediction intervals against the actual complete dataset
fc_naive_drift %>% 
  autoplot(mySeries) +
  labs(title="US Employment in Mining and Logging - Actual vs. Forecast (Seasonal Naive with Drift Model)")

# Calculate point forecast accuracy metrics (MAE, RMSE, MAPE, MASE) on the test set
fc_naive_drift %>% accuracy(mySeries)
```

#### Answering questions to 2nd Benchmark Model

Interpret the **RMSE** (Root Mean Squared Error) in max. 3 sentences:

Interpret the **MAPE** (Mean Absolute Percentage Error) in max. 3 sentences:

## Step 5: Evaluate the Point Forecast Uncertainty of Your 2nd Benchmark Model

```{r}
# Perform residual diagnostics to assess homoscedasticity and normality
naive_drift_employment_fit %>% gg_tsresiduals()

# Extract the upper and lower boundaries for 80% and 95% prediction intervals
fc_naive_drift %>% hilo()
```

#### Answering questions to 2nd Benchmark Model

Are the residuals homoscedastic? (That is, do they have constant variance?)

Are the residuals normally distributed?

Does it make sense to include these prediction intervals in your model evaluation? Why/why not?

# Compare the Evaluation Results of the Two Models
