---
title: "TSAR Project Assignment Part 3 - Group 3"
subtitle: "Times Series Forecasting"
## title-block-banner: true
author:
  - name: Arpad Broglie
    email: arpad.broglie@students.fhnw.ch
    corresponding: true
  - name: Teodor Glisic
    email: teodor.glisic@students.fhnw.ch
    corresponding: true
date: 06/04/2026
date-format: "YYYY"
format: 
  html
editor: visual
execute: 
  warning: false
---

## **Preliminaries**

```{r}
# Install needed libraries to use this report
if (!require(dplyr)) install.packages("dplyr")
if (!require(tidyr)) install.packages("tidyr")
if (!require(ggplot2)) install.packages("ggplot2")
if (!require(fpp3)) install.packages("fpp3")

#Load needed libraries
library(dplyr)
library(ggplot2)
library(tidyr)
library(fpp3)


# Read already filtered time series (tsibble) 
# with Series ID CEU1000000001 (Mining and Logging)
mySeries <- readRDS("us_employment_filtered.rds")
```

## Overview of tsibble

```{r}
head(mySeries)
str(mySeries)
summary(mySeries)
```

## Step 1: Extracting the Training Set

```{r}
# Extracting number of rows for training set with 80/20 split
overall_count <- (nrow(mySeries))
training_row_count <- floor((nrow(mySeries) * 0.8))
train <- mySeries %>% slice(1:training_row_count)
```

## Step 2: Train 1st Benchmark Model

```{r}
# Mean method used
mean_employment_fit <- train %>%
  model(
  Mean = MEAN(Employed)
)

# View first impression of the fitted values
augment(mean_employment_fit)
```

## Step 3: Evaluate the Model Fit of 1st Benchmark Model

### Task A

```{r}
augment(mean_employment_fit) %>% 
  autoplot(.fitted, colour="blue", linetype="dashed") +
  autolayer(train, Employed) +
  labs(y= "US Employment", title="Fitted and true values for monthly US employment rates (Mean method)")
```

#### Answering questions to 1st Benchmark Model

Are the fitted values close to the actual values?

Are they systematically too high or too low?

If yes, where does it come from? (Consider how your benchmark model works!)

### Task B

```{r}
augment(mean_employment_fit) %>%
  ggplot(aes(x=Employed, y=.fitted)) + 
  geom_point() + 
  geom_abline(slope = 1, intercept = 0, linetype=2) + 
  labs(title = "Actual vs. Fitted (Mean Model)", x="Actual", y="Fitted")
           
```

#### Answering questions to 1st Benchmark Model

Are the points close to the identity line?

Are they systematically too high or too low?

If yes, where does it come from? (Relate your answer to what you saw in the time plot.)

### Task C

```{r}
mean_employment_fit %>% gg_tsresiduals()
```

#### Answering questions to 1st Benchmark Model

Are the residuals auto-correlated? How do you decide that based on the plots?

Do the residuals have zero mean? How do you decide that based on the plots?

What do these results tell you about your model?

### Task D

```{r}
augment(mean_employment_fit) %>% 
  features(.innov, box_pierce, lag=10)
```
